---
title: "Take-Home Exercise 4"
description: |
  Putting Visual Analytics into Practical Use.
author:
  - name: Ong Zhi Rong Jordan
    url: https://example.com/norajones
    affiliation: Spacely Sprockets
    affiliation_url: https://example.com/spacelysprokets
date: "`r Sys.Date()`"
output: distill::distill_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = TRUE,
                      include = TRUE,
                      warning = FALSE,
                      message = FALSE)

# Learn more about creating websites with Distill at:
# https://rstudio.github.io/distill/website.html

```

## 1. Overview

In this Exercise, we will reveal the daily routine of **two** randomly selected participant of the city of Engagement, Ohio USA. I will be exploring ViSIElse to visualise their routines.

## 2. Data Preparation

### 2.1 Data Packages

The following packages will be installed to provide several function for data preparation and visualisation.

 - tidyverse: A package with multiple packages that provide the user functions for wrangling, visualising and modelling.
 - ViSiElse: A Visual Tool for Behavior Analysis over Time.
 - lubridate: A package that intuitively changes the date-time object to the user preference.

```{r}
packages = c('tidyverse', 'ViSiElse', 'lubridate')

for(p in packages){
  if(!require(p, character.only = T)){
    install.packages(p)
  }
  library(p, character.only = T)
}
```


### 2.2 Importing of Dataset

Since we are observing the daily routine of the participant, data from **5** status log will be retrieved to observed the participants routine in a **1** month timeframe. To read and combine all the 5 CSVs into one data frame, we first create a function to read and merge the data. We then subseqeuntly call the function with the path of your CSV file.

``` {r}
multmerge = function(mypath){
  filenames=list.files(path=mypath, full.names=TRUE)
  datalist = lapply(filenames, function(x){read.csv(file=x,header=T)})
  Reduce(function(x,y) {merge(x,y,all = TRUE)}, datalist)
}

full_data = multmerge("rawdata/")

```

### 2.3 Data Wrangling and Exploration

#### Retrieving the specific ID from the Data Frame

To retrieve only participant ID **1** and **300** , I first create a data frame to consist of the two numbers. I use the semi_join function to locate the ID of 1 and 300 to form a new data frame, subsequently using the select function to retrieve only the `timestamp`, `participantID`, `currentMode`, `hungerStatus` and `sleepstatus` for this analysis.

``` {r, eval = FALSE}

FilterDF <- data.frame(participantId = c(1, 300), stringsAsFactors = FALSE)

full_data <- semi_join(full_data, FilterDF, by = "participantId")

full_data <- full_data %>%
  select(c("timestamp", "participantId", "currentMode", "hungerStatus","sleepStatus"))

```

#### Splitting the Data Frame
Subsequently, I split the data frame into two different data frame based on their ID to facilitate the creation of the ViSIElse book.

``` {r}
participant1_data <- full_data[full_data$participantId == 1, ]
participant300_data <- full_data[full_data$participantId == 300, ]

```

``` {r, include = FALSE}

participant1_data <- readRDS("data/participant1_data.rds")
participant300_data <- readRDS("data/participant300_data.rds")
```

#### Inserting a new column

Since the ViSIElse book require the time to be in minutes format and an Id, using the `lubridate` package, I converted the timestamp into Year-Mon-Date-Hour-Minutes-Seconds format. I then create another column call `Minutes` using the mutate function and call the `hour`, `minute` and `day` function from `lubridate` for the creation of the two new column `Minutes` and `Day`.

```{r}

Res <- ymd_hms(participant1_data$timestamp)

participant1_data <- participant1_data %>%
  mutate (Minutes = hour(Res)*60 + minute(Res), Day = day(Res))

participant1_data <- participant1_data %>%
  select(c("Day","Minutes", "currentMode", "sleepStatus", "hungerStatus" )) %>%
  rename ( "Current_Mode" = "currentMode",
           "Sleep_Status" = "sleepStatus"
  )


```

#### Generating the Final_Status

This is the tricky part, to generate the status of the participant, we need to identify when the participant start the activity and when the activity stop. We first construct a new column for all the required activities using `cumsum`. This function will construct an increment in value if it did not fulfil the condition inside. 


``` {r}
participant1_data_2 <- participant1_data %>%
  mutate(Sleeping = cumsum(Sleep_Status != "Sleeping")) %>%
  mutate(Working = cumsum(Current_Mode != "AtWork")) %>%
  mutate (Recreation = cumsum(Current_Mode != "AtRecreation")) %>%
  mutate (Hunger = cumsum(hungerStatus != "Hungry"))
```


From the previously created column, we will now identify the change in status using the `lead` and `lag` function. If the current value is different from the subsequent value (using the `lead` function), there is a change in status of your activity which will initate a stop of that activity. Similar to that, if my current value is different from the previous value, it initate a start of a new activity using the `lag` function. This will generate the required labels to identify the change in status of the activities.

``` {r}
participant1_data_3 <- participant1_data_2 %>%
  mutate(Sleep_label = ifelse(Sleeping != lead(Sleeping, default = 0), "Stop_Sleeping", 
                              ifelse(Sleeping == lag(Sleeping, default = 0), "Start_Sleeping",""))) %>% 
  mutate(Work_label = ifelse(Working != lead(Working, default = 0), "Stop_Working", 
                             ifelse(Working == lag(Working, default = 0), "Start_Working",""))) %>% 
  mutate(Recreation_label = ifelse(Recreation != lead(Recreation, default = 0), "Stop_Recreation",
                                   ifelse(Recreation == lag(Recreation, default = 0), "Start_Recreation",""))) %>%
  mutate(Hunger_label = ifelse(Hunger != lead(Hunger, default = 0), "Start_Meal", "")) 

```

Lastly, we will now check for the change in labels to finalise the start and stop of an activity, the period of the activity will be converted to NA.

``` {r}
participant1_data_4 <- participant1_data_3 %>%
  mutate(sleep_status = ifelse(Sleep_label == lag(Sleep_label, default = "NA"), NA , Sleep_label)) %>%
  mutate(work_status = ifelse(Work_label == lag(Work_label, default = "NA"), NA , Work_label)) %>%
  mutate(recreation_status = ifelse(Recreation_label == lag(Recreation_label, default = "NA"), NA, Recreation_label)) %>%
  mutate(hunger_status = ifelse(Hunger_label == lag(Hunger_label, default = "NA"), NA , Hunger_label))


```

``` {r}
supper <- 0:359
breakfast <- 360:720
lunch <- 721:1080
dinner <- 1081:1435
```

``` {r}

participant1_data_4 <- participant1_data_4 %>%
  mutate (final_meal_status = case_when(
    Minutes %in% supper & (hunger_status == 'Start_Meal') & (Sleep_Status != 'Sleeping')  ~ "Supper",
    Minutes %in% breakfast & (hunger_status == 'Start_Meal') & (Sleep_Status != 'Sleeping') ~ "Breakfast",
    Minutes %in% lunch & (hunger_status == 'Start_Meal') & (Sleep_Status != 'Sleeping') ~ "Lunch",
    Minutes %in% dinner & (hunger_status == 'Start_Meal') & (Sleep_Status != 'Sleeping') ~ "Dinner")
  )

```

``` {r}

participant1_data_4$work_status[1] = NA
participant1_data_4$recreation_status[1] = NA

participant1_data_4 <- participant1_data_4 %>%
  select(c("Day", "Minutes", "sleep_status", "work_status",  "recreation_status","final_meal_status"))
```


``` {r}
participant1_data_4 [participant1_data_4 ==""] <- NA

final_data <- participant1_data_4 %>% unite ("Final_Status", sleep_status:final_meal_status, na.rm = T, remove = TRUE)
final_data [final_data ==""] <- NA

final_data <- final_data %>%
  drop_na(Final_Status)

```


``` {r}
pivot_table_1 <- final_data %>%
  pivot_wider(names_from = Final_Status, values_from = Minutes)


pivot_table_1$Start_Working <- sapply(pivot_table_1$Start_Working, head, 1)
pivot_table_1$Stop_Working <- sapply(pivot_table_1$Stop_Working, tail, 1)
pivot_table_1$Start_Recreation <- sapply(pivot_table_1$Start_Recreation, tail, 1)
pivot_table_1$Stop_Recreation <- sapply(pivot_table_1$Stop_Recreation, tail, 1)

colnames(pivot_table_1)[1] <- "id"
```


``` {r}
pivot_table_1[] <- lapply(pivot_table_1, function(x) {
    unlist(ifelse(lengths(x) == 1, x, list(median(unlist(x)))))})

pivot_table_1[] <- lapply(pivot_table_1, function(x) {
    as.integer(x)})

pivot_table_1
```

```{r}
x <- visielse(pivot_table_1)
x
```

``` {r}
x <- visielse(pivot_table_1)
x1 <- ConvertFromViSibook(x@book)
x1
```


``` {r eval=FALSE}
x1 <- x1[order(as.numeric(x1$showorder)), ]
x1$label <- c("Sleep", "Awake", "Work", "Stop Work", "Breakfast", "Have Fun", "Stop Fun", "Dinner")
x1[9,] <- c("sleep", "Sleeping", "l", 1, "Start_Sleeping", "Stop_Sleeping")
x1[10,] <- c ("work", "Working", "l", 3, "Start_Working", "Stop_Working")
x1[11,] <- c ("recreation", "Recreation Time", "l", 4, "Start_Recreation", "Stop_Recreation")
x1$showorder <- c(NA, NA, NA, NA, 2, NA , NA ,5 ,1 ,3 ,4) 
x1 <- x1[order(as.numeric(x1$showorder)), ]
x1
```


``` {r}
v2 <- visielse(pivot_table_1, 
               book = x1, 
               informer = NULL, 
               doplot = F, 
               pixel = 30)

plot(v2, 
     vp0w = 0.7, 
     unit.tps = "min", 
     scal.unit.tps = 30, 
     main = "Typical day")
```

